---
title: "R vs. Julia LMEM comparisons"
author: "June Choe"
date: "2022-08-25"
format:
  html:
    theme: flatly
    toc: true
    page-layout: full
knitr:
  opts_chunk:
    dev: "svglite"
    dev.args:
      scaling: 1.5
filters:
  - grouped-tabsets
cache: true
---

## Setup

::: {.panel-tabset group="language"}

## R

```{r R-setup}
#| message: false
library(arrow)
library(lme4); library(lmerTest)
library(broom); library(broom.mixed)
```

Extra setup to use Julia in Quarto:

```{r R-setup-misc}
#| message: false
library(JuliaCall)
system.time( julia_setup() )
```

## Julia

Libraries:

```{julia J-setup}
using MKL, Arrow, DataFrames, MixedModels, StatsModels
```

Extra setup for interactive REPL:

```{julia J-setup-misc}
#| results: hide
using ProgressMeter, InteractiveUtils
ProgressMeter.ijulia_behavior(:clear)
```

:::

## 1) Linear regression predicting AY vheight

### Data from **PNC**

::: {.panel-tabset group="language"}

## R

```{r R-1-data}
PNC_ay <- read_feather("PNC_ay.arrow")
```

```{r R-1-show-data}
summary(PNC_ay)
```

## Julia

```{julia J-1-data}
PNC_ay = Arrow.Table("PNC_ay.arrow")
```

```{julia J-1-show-data}
describe(DataFrame(PNC_ay))
```

:::

### Model 1: Maximal

::: {.panel-tabset group="language"}

## R

R dies 😤

Not run:

```{r R-1-m1}
#| eval: false
system.time(
  R_lmm_1 <- lmer(
    vheight ~ birthyear_z2 * allophone * gender + logdur_z2 + frequency_z2 +
                  (1 + allophone + logdur_z2 + frequency_z2 | participant) +
                  (1 + birthyear_z2 * gender + logdur_z2 | word),
    data = PNC_ay
  )
)
```

## Julia

Julia goes brrr

```{julia J-1-m1}
J_lmm_1 = @time fit(
  MixedModel,
  @formula(
    vheight ~ birthyear_z2 * allophone * gender + logdur_z2 + frequency_z2 +
                  (1 + allophone + logdur_z2 + frequency_z2 | participant) +
                  (1 + birthyear_z2 * gender + logdur_z2 | word)
  ),
  PNC_ay
)
```

No singular fit!

```{julia J-1-m1-singular}
issingular(J_lmm_1)
```

:::

### Model 2: Zero correlation

::: {.panel-tabset group="language"}

## R

R dies 😤

Not run:

```{r R-1-m2}
#| eval: false
system.time(
  R_lmm_2 <- lmer(
    vheight ~ birthyear_z2 * allophone * gender + logdur_z2 + frequency_z2 +
                  (1 + allophone + logdur_z2 + frequency_z2 || participant) +
                  (1 + birthyear_z2 * gender + logdur_z2 || word),
    data = PNC_ay
  )
)
```

## Julia

Still fits, now faster

```{julia J-1-m2}
J_lmm_2 = @time fit(
  MixedModel,
  @formula(
    vheight ~ birthyear_z2 * allophone * gender + logdur_z2 + frequency_z2 +
                  zerocorr(1 + allophone + logdur_z2 + frequency_z2 | participant) +
                  zerocorr(1 + birthyear_z2 * gender + logdur_z2 | word)
  ),
  PNC_ay
)
```

Best to keep it maximal, though

```{julia J-1-m1-m2-anova}
MixedModels.likelihoodratiotest(J_lmm_1, J_lmm_2)
```

:::

### Model 3: Intercept Only

::: {.panel-tabset group="language"}

## R

Finally fits

```{r R-1-m3}
system.time(
  R_lmm_3 <- lmer(
    vheight ~ birthyear_z2 * allophone * gender + logdur_z2 + frequency_z2 +
                  (1 | participant) +
                  (1 | word),
    data = PNC_ay
  )
)
```

```{r R-1-m3-ranef}
tidy(R_lmm_3, effects = "ran_pars")
```

```{r R-1-m3-fixed}
tidy(R_lmm_3, effects = "fixed")
```

## Julia

brrr

```{julia J-1-m3}
J_lmm_3 = @time fit(
  MixedModel,
  @formula(
    vheight ~ birthyear_z2 * allophone * gender + logdur_z2 + frequency_z2 +
                  (1 | participant) +
                  (1 | word)
  ),
  PNC_ay
)
```

:::


### Misc: Bringing Julia model to R

Load libraries for interfacing with R/lmer:

```{julia jellyme4}
#| eval: false
using JellyMe4, RCall
```

Pack original df with the model object:

```{julia augment-model}
#| eval: false
J_lmm_1_packed = (J_lmm_1, DataFrame(PNC_ay))
```

Send to R:

```{julia transfer-to-R}
#| eval: false
@rput J_lmm_1_packed
```

Save model object:

```{julia write-transferred-mod}
#| eval: false
R"saveRDS(J_lmm_1_packed, 'J_lmm_1_packed.rds')"
```

Read it back in R:

```{r read-transferred-mod}
J_lmm_1_inR <- readRDS("J_lmm_1_packed.rds")
tidy(J_lmm_1_inR, effects = "fixed")
```

Visualize the convergence of a random effect (word intercept) with ggplot:

```{r viz-convergence}
#| message: false
#| warning: false
# Extract info from model
theta <- getME(J_lmm_1_inR, "theta")
word_intercept_theta <- theta[1]
fit_fn <- as.function(J_lmm_1_inR)
fit_fn_marginalized <- function(word_intercept) {
  fit_fn(replace(theta, 1, word_intercept))
}
# Plot
library(dplyr)
library(ggplot2)
tibble(
  x = seq(0, 2, length = 21),
  y = sapply(x, fit_fn_marginalized)
) %>% 
  ggplot(aes(x, y)) +
  geom_line() +
  annotate("point", size = 2, color = "red",
           x = word_intercept_theta,
           y = fit_fn_marginalized(word_intercept_theta)) +
  scale_y_continuous(expand = expansion(.1)) +
  labs(
    title = "(Marginalized) optimization of by-word intercept",
    y = "Deviance (a.k.a. Objective)",
    x = "Scaled standard deviation"
  )
```


### Misc: re-fitting Julia maximal model with REML

```{julia J-1-m1-reml}
@time refit!(J_lmm_1, REML = true)
```

<br>
<br>

## 2) Logistic regression predicting accuracy

### Data from **Choe, Yoshida, & Cole (2022)**

::: {.panel-tabset group="language"}

## R

```{r R-2-data}
CYC_2022 <- read_feather("CYC_2022.arrow")
```

```{r R-2-show-data}
#| echo: false
rmarkdown::paged_table(CYC_2022)
```

## Julia

```{julia J-2-data}
CYC_2022 = Arrow.Table("CYC_2022.arrow")
```

:::

### Coding scheme

::: {.panel-tabset group="language"}

## R

Explicit sum-coding of `Condition`

```{r R-2-coding}
CYC_2022$Condition <- c("Verb" = 1, "Subject" = -1)[CYC_2022$Condition]
```

## Julia

Coding dictionary with sum-coding of `Condition` and groups marked with `Grouping()`

```{julia J-2-coding}
#| results: hide
coding2 = Dict(
  :Subject => Grouping(),
  :Item => Grouping(),
  :Condition => EffectsCoding(base = "Subject")
)
```

:::

### Model 1: Maximal

::: {.panel-tabset group="language"}

## R

R dies 😤

```{r R-2-m1}
#| eval: false
R_lmm_1 <- lmer(
  Accuracy ~ Condition * SemanticFit * Transitivity +
    (1 + Condition | Item) +
    (1 + Condition * SemanticFit * Transitivity | Subject),
  data = CYC_2022, family = binomial()
)
```

## Julia

Converges...

```{julia J-2-m1}
J_glmm_1 = @time fit(
  MixedModel,
  @formula(
    Accuracy ~ Condition * SemanticFit * Transitivity +
        (1 + Condition * SemanticFit * Transitivity | Subject) +
        (1 + Condition | Item)
  ),
  CYC_2022, Binomial(), contrasts = coding2
)
```

... but with singular fit

```{julia J-2-m1-singular}
issingular(J_glmm_1)
```

:::

### Model 2: Zero correlation

::: {.panel-tabset group="language"}

## R

R dies again 😤

```{r R-2-m2}
#| eval: false
R_glmm_2 <- glmer(
  Accuracy ~ Condition * SemanticFit * Transitivity +
    (1 + Condition || Item) +
    (1 + Condition * SemanticFit * Transitivity || Subject),
  data = CYC_2022, family = binomial()
)
```

## Julia

Converges...

```{julia J-2-m2}
J_glmm_2 = @time fit(
  MixedModel,
  @formula(
    Accuracy ~ Condition * SemanticFit * Transitivity +
        zerocorr(1 + Condition * SemanticFit * Transitivity | Subject) +
        zerocorr(1 + Condition | Item)
  ),
  CYC_2022, Binomial(), contrasts = coding2
)
```

... butwith singular fit, again

```{julia J-2-m2-singular}
issingular(J_glmm_2)
```

PCA on random effects structure:

```{julia J-2-m2-repca}
J_glmm_2.rePCA
```

:::


### Model 3: Zero variance components dropped

::: {.panel-tabset group="language"}

## R

R doesn't die, but model fails to converge

```{r R-2-m3}
system.time(
  R_glmm_3 <- glmer(
    Accuracy ~ Condition * SemanticFit * Transitivity +
      (1 | Item) +
      (1 + Condition | Subject),
    data = CYC_2022, family = binomial()
  )
)
```

## Julia

Converges ...

```{julia J-2-m3}
J_glmm_3 = @time fit(
  MixedModel,
  @formula(
    Accuracy ~ Condition * SemanticFit * Transitivity +
        (1 + Condition | Subject) +
        (1 | Item)
  ),
  CYC_2022, Binomial(), contrasts = coding2
)
```

... no singular fit!

```{julia J-2-m3-singular}
issingular(J_glmm_3)
```

:::

### Model 4: Reduced - some interaction terms dropped

::: {.panel-tabset group="language"}

## R

Converges

```{r R-2-m4}
system.time(
  R_glmm_4 <- glmer(
    Accuracy ~ Condition + SemanticFit * Transitivity +
      (1 | Item) +
      (1 + Condition | Subject),
    data = CYC_2022, family = binomial()
  )
)
```

```{r R-2-m4-ranef}
tidy(R_glmm_4, effects = "ran_pars")
```

```{r R-2-m4-coef}
tidy(R_glmm_4, effects = "fixed")
```

## Julia

Converges, of course

```{julia J-2-m4}
J_glmm_4 = @time fit(
  MixedModel,
  @formula(
    Accuracy ~ Condition + SemanticFit * Transitivity +
        (1 + Condition | Subject) +
        (1 | Item)
  ),
  CYC_2022, Binomial(), contrasts = coding2
)
```

:::

### Model comparison

Reduced model is actually adequate, and you might prefer this if you care about parsimony, but only Julia can give us this info:

```{julia J-2-m3-m4-anova}
MixedModels.likelihoodratiotest(J_glmm_3, J_glmm_4)
```

R actually can give us the same info, but only because the max model that failed to converge happened to get pretty close.

```{r R-2-m3-m4-anova}
anova(R_glmm_3, R_glmm_4)
```

Generally you can't do tests like this on models that fail to converge. While the convergence issue here was just one of performance (lower threshold or # of iterations would've sufficed), you can't know this a priori. You also cannot make any claims about how trivial the contribution of the dropped interaction effects were in the `R_glmm_3` model because it failed to converge.

### Misc: `fast = true` fitting of maximal model

Only available for GLMM:

```{julia J-2-m1-fast}
@time fit(
  MixedModel,
  @formula(
    Accuracy ~ Condition * SemanticFit * Transitivity +
        (1 + Condition * SemanticFit * Transitivity | Subject) +
        (1 + Condition | Item)
  ),
  CYC_2022, Binomial(), contrasts = coding2,
  fast = true
)
```

<br>
<br>

## Version info

::: {.panel-tabset group="language"}

## R

```{r R-session}
R.version
packageVersion("lme4")
```

## Julia

```{julia J-session}
versioninfo()
using Pkg
Pkg.status("MixedModels")
```

:::
